### 2021-01-26 paper/arm.tex
Extra polish

\begin{example}
  Revisiting \refex{ex:rrd1} and eliding irrelevant transformers:
  \begin{align*}
    \begin{gathered}[t]
      \PR{y}{r}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
          \event{b}{\DR{y}{1}}{}
          \xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
        \end{tikzinline}}  
    \end{gathered}  
    &&
    \begin{gathered}[t]
      \IF{r}\THEN s\GETS x\FI
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{c}{r\mid\DW{x}{0}}{}
          % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
          % \xform{x1i}{\lnot\Qw{y}\land\bForm[\TRUE/\Dx{y}]}{below=of x1d}
          % \xo{c}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  Composing sequentially:
  \begin{align*}
    \begin{gathered}[t]
      \PR{y}{r}
      \SEMI
      \IF{r}\THEN s\GETS x\FI
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{b}{\DR{y}{1}}{}
          \event{c}{\RW\limplies r\mid\DW{x}{0}}{right=of b}
          %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  Embedding the thread in thread group, \ref{T3} yields:
  \begin{align*}
    \begin{gathered}[t]
      % \PR{y}{r}
      % \SEMI
      % \IF{r}\THEN s\GETS x\FI
      % \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{b}{\DR{y}{1}}{}
          \event{c}{\FALSE\limplies r\mid\DW{x}{0}}{right=of b}
          %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  The precondition of $\DWP{x}{0}$ is a tautology, as required.
\end{example}

@@

\begin{definition}[$\xCO$/$\xRASC$/$\xDGR$/$\xRRD$]
  \label{def:pomsets-arm}
  Update \refdef{def:pomsets-trans} to (\ref{L4} unchanged):
  \begin{enumerate}
  \item[\ref{S3})]
    $\labelingForm(\aEv)$ implies $\QS{\aLoc}{\amode}\land\aExp{=}\aVal$,
  \item[\ref{L3})]
    $\labelingForm(\aEv)$ implies $\QL{\aLoc}{\amode}$.
  \item[\ref{T3})]
    $\labelingForm(\aEv)$ implies
    $\labelingForm_1(\aEv)[\TRUE/\Q{}][\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
    \\
    $\labelingForm(\aEv)$ implies
    $\labelingForm_1(\aEv)[\TRUE/\Q{}][\FALSE/\RW]$ otherwise.
  \end{enumerate}
  \begin{enumerate}
  \item[\ref{S4})]
    $\aTr{\bEvs}{\bForm}$ implies
    \begin{math}
      %(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land
      \bForm %\noSUB{[\aExp/\aLoc]},
      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]
      \DS{\aLoc}{\amode}
    \end{math}
  \item[\ref{S5})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      %\lnot\Qw{\aLoc}\land 
      \bForm %\noSUB{[\aExp/\aLoc]}.
      [\FALSE/\QS{\aLoc}{\amode}]
      \DS{\aLoc}{\amode}
    \end{math}
  \item[\ref{L4})]
    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
  \item[\ref{L5})]
    % $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land(\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
    % \item[\ref{L6})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      % \lnot\Qx{\aLoc} \land
      \DL{\aLoc}{\amode}{} \land (\RW\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]).
    \end{math}
  \end{enumerate}  
\end{definition}

>>>>

\begin{example}
  Revisiting \refex{ex:rrd1} and eliding irrelevant transformers:
  \begin{align*}
    \begin{gathered}[t]
      \PR{y}{r}
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and .5em]
          \event{b}{\DR{y}{1}}{}
          \xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
        \end{tikzinline}}  
    \end{gathered}  
    &&
    \begin{gathered}[t]
      \IF{r}\THEN s\GETS x\FI
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{c}{r\mid\DR{x}{0}}{}
          % \xform{x1d}{\bForm[\TRUE/\Dx{\aLoc}]}{below=of c}
          % \xform{x1i}{\lnot\Qw{y}\land\bForm[\TRUE/\Dx{y}]}{below=of x1d}
          % \xo{c}{x1d}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  Composing sequentially:
  \begin{align*}
    \begin{gathered}[t]
      \PR{y}{r}
      \SEMI
      \IF{r}\THEN s\GETS x\FI
      \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{b}{\DR{y}{1}}{}
          \event{c}{\RW\limplies r\mid\DR{x}{0}}{right=of b}
          %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  Embedding the thread in thread group, \ref{T3} yields:
  \begin{align*}
    \begin{gathered}[t]
      % \PR{y}{r}
      % \SEMI
      % \IF{r}\THEN s\GETS x\FI
      % \\
      \hbox{\begin{tikzinline}[node distance=0.5em and 1.5em]
          \event{b}{\DR{y}{1}}{}
          \event{c}{\FALSE\limplies r\mid\DR{x}{0}}{right=of b}
          %\xform{x1i}{\RW\limplies\bForm}{right=of b} %x1d}
        \end{tikzinline}}  
    \end{gathered}  
  \end{align*}
  The precondition of $\DRP{x}{0}$ is a tautology, as required.
\end{example}

@@

\begin{definition}[$\xCO$/$\xRASC$/$\xDGR$/$\xRRD$]
  \label{def:pomsets-arm}
  Update \refdef{def:pomsets-trans} to: % (\ref{L4} unchanged):
  \begin{enumerate}
  \item[\ref{S3})]
    $\labelingForm(\aEv)$ implies $\QS{\aLoc}{\amode}\land\aExp{=}\aVal$,
  \item[\ref{L3})]
    $\labelingForm(\aEv)$ implies $\QL{\aLoc}{\amode}$.
  \item[\ref{T3})]
    $\labelingForm(\aEv)$ implies
    $\labelingForm_1(\aEv)[\TRUE/\Q{}][\TRUE/\RW]$ if $\labelingAct_1(\aEv)$ is a write,
    \\
    $\labelingForm(\aEv)$ implies
    $\labelingForm_1(\aEv)[\TRUE/\Q{}][\FALSE/\RW]$ otherwise.
  \end{enumerate}
  \begin{enumerate}
  \item[\ref{S4})]
    $\aTr{\bEvs}{\bForm}$ implies
    \begin{math}
      %(\Qw{\aLoc}\limplies\aExp{\neq}\aVal)\land
      \bForm %\noSUB{[\aExp/\aLoc]},
      [(\Qw{\aLoc}\land\aExp{=}\aVal)/\Qw{\aLoc}]
      \DS{\aLoc}{\amode}
    \end{math}
  \item[\ref{S5})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      %\lnot\Qw{\aLoc}\land 
      \bForm %\noSUB{[\aExp/\aLoc]}.
      [\FALSE/\QS{\aLoc}{\amode}]
      \DS{\aLoc}{\amode}
    \end{math}
  \item[\ref{L4})]
    $\aTr{\bEvs}{\bForm}$ implies $\aVal{=}\aReg\limplies\bForm$, 
  \item[\ref{L5})]
    % $\aTr{\cEvs}{\bForm}$ implies $\lnot\Qx{\aLoc} \land \DL{\aLoc}{\amode} \land(\RW\limplies(\aVal{=}\aReg\lor\aLoc{=}\aReg)\limplies\bForm)$,
    % \item[\ref{L6})]
    $\aTr{\cEvs}{\bForm}$ implies
    \begin{math}
      % \lnot\Qx{\aLoc} \land
      \DL{\aLoc}{\amode}{} \land (\RW\limplies\bForm[\FALSE/\QL{\aLoc}{\amode}]).
    \end{math}
  \end{enumerate}  
\end{definition}

### 2021-01-26 paper/intro.tex
Extra polish

This model naturally extends to the case of shared-memory concurrency
in a natural way, leading to a \emph{sequentially consistent}
semantics, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=20,in=160]{wx1}{rx1}
        \po{wx0}{wx1}
        \po{wx1}{wy2}
        \po{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}

Unfortunately, this model does not compile efficiently to commodity
hardware, resulting in a 37--73\% increase in CPU time~\cite{Liu:2019:ASC:3314221.3314611} on ARM, and
hence power consumption.  Developers of software and compilers have
therefore been faced with a difficult trade-off, between an elegant
model of memory, and its impact on resource usage (such as size of
data centers, electricity bills and carbon footprint). Unsurprisingly,
many have chosen to prioritize efficiency over elegance.

@@

For example, for single-event pomsets, there are two predicate
transformers, since there are two subsets of any one-element set.
We call the predicate transformer for $\emptyset$ the \emph{independent}
transformer, and the one indexed by $\{e\}$ the \emph{dependent} transformer.
We visualize this by including more than one transformed predicate,
with an edge leading to the dependent one. For example:
  \begin{align*}
    \begin{gathered}
      r \GETS x
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
          \event{rx0}{\DR{x}{0}}{}
          \xform{rx0i}{\bForm}{above=of rx0}
          \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
          \xo{rx0}{rx0d}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      s \GETS y
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
          \event{ry0}{\DR{y}{0}}{}
          \xform{ry0i}{\bForm}{above=of ry0}
          \xform{ry0d}{(s{=}0) \limplies \bForm}{below=of ry0}
          \xo{ry0}{ry0d}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      \IF{s{<}1} \THEN z\GETS r{*}s \FI
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
          \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
          \xform{wz0i}{\bForm}{above=of wz0}
          \xform{wz0d}{\bForm}{below=of wz0}
          \xo{wz0}{wz0d}
      \end{tikzinline}}
    \end{gathered}
  \end{align*}
The model of sequential composition then picks which
predicate transformer to apply to an event's precondition by picking
the one indexed by all the events before it in causal order.

For example, we can recover the expected semantics
for the above example by choosing
the predicate transformer which is independent of $(\DR x0)$
but dependent on $(\DR y0)$, which is the transformer
which maps $\bForm$ to $(s{=}0) \limplies \bForm$.
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
        \xform{rx0d}{(r{=}0) \limplies \bForm}{above=of rx0}
        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
        \xform{wz0i}{\bForm}{above=of wz0}
        \po{ry0}{wz0}
        \xo{rx0}{rx0d}
        \xo{ry0}{ry0d}
        \xo{rx0}{rx0ry0d}
        \xo{ry0}{rx0ry0d}
      \end{tikzinline}}
  \end{gather*}
As a sanity check, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left, with intermediate step:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{below=of ry0}
        \xform{rx0d}{(r{=}0) \limplies \bForm}{above=of rx0}
        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
        \xform{wz0i}{\bForm}{right=of ry0d}
        \xo{rx0}{rx0d}
        \xo{ry0}{ry0d}
        \xo{rx0}{rx0ry0d}
        \xo{ry0}{rx0ry0d}
      \end{tikzinline}}
  \end{gather*}
or to the right, with intermediate step:
    \begin{gather*}
    s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{0}}{}
        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \xform{ry0d}{(s{=}0) \limplies \bForm}{above=of ry0}
        \xform{wz0i}{\bForm}{above=of wz0}
        \po{ry0}{wz0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gather*}
This is an instance of a general result that sequential composition forms a monoid,
as one would hope.

>>>>

This model naturally extends to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=20,in=160]{wx1}{rx1}
        \po{wx0}{wx1}
        \po{wx1}{wy2}
        \po{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}

Unfortunately, this model does not compile efficiently to commodity
hardware, resulting in a 37--73\% increase in CPU time on ARM~\cite{Liu:2019:ASC:3314221.3314611} and,
hence, in power consumption.  Developers of software and compilers have
therefore been faced with a difficult trade-off, between an elegant
model of memory, and its impact on resource usage (such as size of
data centers, electricity bills and carbon footprint). Unsurprisingly,
many have chosen to prioritize efficiency over elegance.

@@

For example, for single-event pomsets, there are two predicate
transformers, since there are two subsets of any one-element set.
The \emph{independent}
transformer is indexed by the empty set, whereas the \emph{dependent}
transformer is indexed by the singleton.
We visualize this by including more than one transformed predicate,
with an edge leading to the dependent one. For example:
  \begin{align*}
    \begin{gathered}
      r \GETS x
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
          \event{rx0}{\DR{x}{0}}{}
          \xform{rx0i}{\bForm}{above=of rx0}
          \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
          \xo{rx0}{rx0d}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      s \GETS y
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
          \event{ry0}{\DR{y}{0}}{}
          \xform{ry0i}{\bForm}{above=of ry0}
          \xform{ry0d}{(s{=}0) \limplies \bForm}{below=of ry0}
          \xo{ry0}{ry0d}
        \end{tikzinline}}
    \end{gathered}
    &&
    \begin{gathered}
      \IF{s{<}1} \THEN z\GETS r{*}s \FI
      \\
      \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
          \event{wz0}{(s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{}
          \xform{wz0i}{\bForm}{above=of wz0}
          \xform{wz0d}{\bForm}{below=of wz0}
          \xo{wz0}{wz0d}
      \end{tikzinline}}
    \end{gathered}
  \end{align*}
The model of sequential composition then picks which
predicate transformer to apply to an event's precondition by picking
the one indexed by all the events before it in causal order.

For example, we can recover the expected semantics
for the above example by choosing
the predicate transformer which is independent of $(\DR x0)$
but dependent on $(\DR y0)$, which is the transformer
which maps $\bForm$ to $(s{=}0) \limplies \bForm$.
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{above=of ry0}
        \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
        \xform{wz0i}{\bForm}{above=of wz0}
        \po{ry0}{wz0}
        \xo{rx0}{rx0d}
        \xo{ry0}{ry0d}
        \xo{rx0}{rx0ry0d}
        \xo{ry0}{rx0ry0d}
      \end{tikzinline}}
  \end{gather*}
As a sanity check, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left, with intermediate step:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \xform{rx0ry0d}{(r{=}0) \limplies(s{=}0) \limplies \bForm}{above=of ry0}
        \xform{rx0d}{(r{=}0) \limplies \bForm}{below=of rx0}
        \xform{ry0d}{(s{=}0) \limplies \bForm}{right=of rx0d}
        \xform{wz0i}{\bForm}{right=of rx0ry0d}
        \xo{rx0}{rx0d}
        \xo{ry0}{ry0d}the \emph{independent}
transformer
        \xo{rx0}{rx0ry0d}
        \xo{ry0}{rx0ry0d}
      \end{tikzinline}}
  \end{gather*}
or to the right, with intermediate step:
    \begin{gather*}
    s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r{*}s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{0}}{}
        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r{*}s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \xform{ry0d}{(s{=}0) \limplies \bForm}{above=of ry0}
        \xform{wz0i}{\bForm}{above=of wz0}
        \po{ry0}{wz0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gather*}
This is an instance of a general result that sequential composition forms a monoid,
as one would hope.
